Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-26 doc/paper/paper.tex
Some writing and some more feedback

% TODO: I found the choice for what goes into a footnote and what doesn't rather
% inconsistent, which made it a bit irritating to read for me. Maybe review if
% the current use of footnotes is really necessary and helpful?
% TODO: anna's feedback

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}

@@

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is some previous work
  showing how FOMC can compute
  factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
  One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
  not directly correspond to any algorithm.
  Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
  liftable but contains no experimental results. The
  third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
  with an algorithm, which is not publicly available.} We claim that the
capabilities of (W)FOMC algorithms can be significantly extended by empowering
them with the ability to construct recursive solutions. The topic of recursion
in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Second, while Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} allow for equality
constraints and constraints of the form $X \not\in \Delta$ for some variable $X$
and domain $\Delta$, we exclude such constraints simply because they are not
needed.

@@

@@

\section{How to Interpret an FCG}\label{sec:interpret}
% TODO: rewrite this section

\begin{algorithm}[t]
  \caption{Construct functions from an FCG}
  \SetKwProg{Fn}{Function}{:}{}

@@

  \KwIn{$\mathcal{D}$---the set of domains of the input formula}
  \KwIn{$s$---the source node of the FCG}
  \KwData{$\functionNames = \emptyset$ (function names)}
  \KwOut{a list of function definitions}
  $\variableNames \gets \{\, \Omega \mapsto \newVariableName{} \mid \Omega \in \mathcal{D} \,\}$\;
  \If{$s$ is a direct successor of a $\Reff$ node}{
    $(\expression, \functions) \gets \visit{$s$, $\variableNames$}$\;
    \Return{\functions}\;
  }
  $f \gets \newFunctionName{$s$}$\;
  $(\expression, \functions) \gets \visit{$s$, $\variableNames$}$\;
  \Return{$\langle f(\langle \variableNames{$\Omega$} \mid \Omega \in \mathcal{D} \rangle) = \expression \rangle \mdoubleplus \functions$}\;
  \Fn{\visit{node $v$, variable names \variableNames}}{
    \If{$v$ is \textbf{not} a direct successor of a $\Reff$ node}{
      \Return{\actuallyVisit{$v$, \variableNames}}\;
    }
    $f \gets \newFunctionName{$v$}$\;
    $\newVariableNames \gets \variableNames$\;
    \For{$\Omega \in \mathfrak{D}(v)$ s.t. $\newVariableNames(\Omega)$ is non-atomic}{
      $\newVariableNames(\Omega) \gets \newVariableName{}$\;
    }
    $(\expression, \functions) \gets \actuallyVisit{$v$, \newVariableNames}$\;
    $\functionCall \gets f(\langle \variableNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    $\functionSignature \gets f(\langle \newVariableNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    \Return{$(\functionCall, \langle \functionSignature = \expression \rangle \mdoubleplus \functions)$}\;
  }
  \Fn{\actuallyVisit{node $v$, variable names \variableNames}}{
    \Switch{label of $v$}{
      \lCase{$\GDR(v')$}{\Return{\visit{$v'$, \variableNames}}}
      \uCase{$\CR_{\Omega \mapsto \Omega'}(v')$}{
        \Return{\visit{$v'$, $\variableNames \cup \{\, \Omega' \mapsto (\variableNames(\Omega) - 1) \,\}$}}\;
      }
      \uCase{$\Reff_{\rho}(v')$}{
        $\args \gets \langle \variableNames{$\rho(\Omega)$} \mid \Omega \in \mathfrak{D}(v') \rangle$\;
        \Return{$(\functionNames{$v'$}(\args), \langle \rangle)$}\;
      }
      \dots
    }
  }
\end{algorithm}

% \begin{itemize}
%   \item atomic --- no subtraction, i.e., just a variable name
%   \item for each FCG node $v$, $\mathfrak{D}(v)$ is a precomputed set of domains
%         that are used by $v$ and all of its successors (except the ones that are
%         introduced by $v$ or one of its successors) (this was already part of
%         \textsc{ForcLift}, although had to be reimplemented in a way that avoids
%         infinite loops when dealing with cycles in graphs)
%   \item for list generators over a set of domains, we assume an arbitrary fixed
%         ordering over all domains
%   \item $\mathsf{F}$ is a partial map from the set of FCG nodes to Strings,
%         i.e., function names
%   \item \newVariableName just generates a unique identifier for a variable
%   \item \newFunctionName does the same thing for function names and also records
%         the new name in $\mathsf{F}$
%   \item the top function is the main one
%   \item `is a direct successor of a $\Reff$ node' --- the in-degree is greater
%         than 1
% \end{itemize}

@@

\begin{example}\label{example:interpretation}
  Consider the FCG from \cref{fig:examplefcg}. The input formula (i.e., the
  formula from \cref{example:first}) has two domains: $\Gamma$ and $\Delta$.
  Thus, the interpretation of the FCG is a function
  $f\colon \mathbb{N}_{0} \times \mathbb{N}_{0} \to \mathbb{R}_{\ge 0}$. Let
  $m \coloneqq |\Gamma|$, and $n \coloneqq |\Delta|$. The node labelled
  $\bigvee_{\Delta^{\top} \subseteq \Delta}$ tells us that
  $f(m, n) = \sum_{l = 0}^{n} \binom{n}{l} \square$, where $\square$ is the
  interpretation of the remaining subgraph, and $l$ iterates over all possible
  sizes of $\Delta^{\top}$. It also creates two subdomains
  $\Delta^{\top}, \Delta^{\bot} \subseteq \Delta$ that partition $\Delta$, i.e.,
  as the size of $\Delta^{\top}$ increases, the size of $\Delta^{\bot}$
  correspondingly decreases. Nodes labelled $\land$ correspond to
  multiplication. Therefore,
  $f(m, n) = \sum_{l = 0}^{n} \binom{n}{l} \diamondsuit \times \heartsuit$,
  where $\diamondsuit$ is the interpretation of the contradiction (i.e., $\bot$)
  node, and $\heartsuit$ is the interpretation of the $\Reff$ node.

  A contradiction node with clause $c$ as a parameter is interpreted as one if
  the clause has groundings and zero otherwise. In this case, the parameter is
  $c = (\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\})$
  (not shown in \cref{fig:examplefcg}), which can be read as
  $\forall X, Y \in \Delta^{\top}\text{. }X \ne Y \Rightarrow \bot$, i.e.,
  $\forall X, Y \in \Delta^{\top}\text{. }X = Y$. This latter sentence is true
  if and only if $|\Delta^{\top}| < 2$. Therefore, we can use the Iverson
  bracket notation to write
  \[
    \diamondsuit = [l < 2] \coloneqq
    \begin{cases}
      1 & \text{if } l < 2 \\
      0 & \text{otherwise.}
    \end{cases}
  \]

  It remains to interpret the $\Reff$ node. Parameter
  $\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^\bot \,\}$ tells us that
  the interpretation of the $\Reff$ node should be the same as that of the
  source node, but with domains $\Gamma$ and $\Delta$ replaced with $\Gamma'$
  and $\Delta^{\bot}$, respectively. Domain $\Gamma'$ was created by the CR rule
  applied on $\Gamma$, so $|\Gamma'| = m - 1$. Now
  $\Delta^{\bot} = \Delta \setminus \Delta^{\top}$, and $|\Delta^{\top}| = l$,
  so $|\Delta^{\bot}| = n - l$. Thus, the interpretation of the $\Reff$ node is
  a recursive call to $f(m - 1, n - l)$. Therefore,
  \begin{align}
    f(m, n) &= \sum_{l = 0}^{n} \binom{n}{l} [l < 2] f(m-1, n-l)\nonumber \\
            &= f(m-1, n) + n f(m-1, n-1).\label{eq:solution}
  \end{align}
  To use this recursive function to compute the model count of the input formula
  for any domain sizes, one just needs to find the base cases $f(0, n)$ and
  $f(m, 0)$ for all $m, n \in \mathbb{N}_{0}$.
\end{example}

\section{Experimental Results}\label{sec:results}

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}

@@

% how the complexity of a solution is determined
Previous work often compares WFOMC algorithms by running them on a few instances
with increasing domain sizes and measuring
runtime~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/aaai/BroeckD12}.
However, we can do much better and identify the exact worst-case asymptotic
complexity of a solution produced by either \textsc{Crane} or \textsc{ForcLift}.
The asymptotic complexity of a solution can be determined by considering the
total number of arithmetic operations needed to follow the definitions of
constituent functions without recomputing the same quantity multiple times. In
particular, we assume that each function call and binomial coefficient is
computed at most once. We also assume that computing $\binom{n}{k}$ takes
$\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on the computation of each $f(i, j)$.

>>>>

% TODO: I found the choice for what goes into a footnote and what doesn't rather
% inconsistent, which made it a bit irritating to read for me. Maybe review if
% the current use of footnotes is really necessary and helpful?
% TODO: replace references with the condensed version
% TODO: feedback2.pdf
% TODO: feedback1.pdf

% TODO: what's the punchline? make sure the introduction tells the reader what
% to expect from section 5

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Vone}{\mathsf{V}_{1}}
\newcommand{\Vtwo}{\mathsf{V}_{2}}
\newcommand{\Vthree}{\mathsf{V}_{3}}

@@

% TODO: also cite 'WFOMC with DAG axioms'
% \cite{DBLP:journals/corr/abs-2302-09830}

% TODO: add a flow graph: \phi -> Box (Compilation) -> \G_\phi -> Box (Algorithm
% 4) -> \mathcal{F}_\phi -> Box (simplification) -> \mathcal{F|_\phi, base
% cases, domain sizes -> Box (Propagation (?))

% TODO: too many acronyms (e.g., FOL)

@@

% TODO: must remove this footnote
However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is some previous work
  showing how FOMC can compute
  factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
  One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
  not directly correspond to any algorithm.
  Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
  liftable but contains no experimental results. The
  third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
  with an algorithm, which is not publicly available.} We claim that the
capabilities of (W)FOMC algorithms can be significantly extended by empowering
them with the ability to construct recursive solutions. The topic of recursion
in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Second, while Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} allow for equality
constraints and constraints of the form $X \not\in \Delta$ for some variable $X$
and domain $\Delta$, we exclude such constraints simply because they are not
needed.
% TODO: explain why we're not losing anything by dropping equality constraints
% (e.g., X = Y). Answer: having two variables that are equal is the same thing
% as having one variable (since by the definition of a clause, this equality can
% only be at the beginning of a clause.

@@

% TODO: the definition of FCG is hard to miss (highlight its importance, maybe
% remind the reader later on in the paper)

% TODO: instead of FO d-DNNF computational graph, say 'first-order computational
% graph' (FOCG?)

@@

\section{Converting FCGs to Function Definitions}\label{sec:interpret}

\begin{algorithm}[t]
  \caption{Construct functions from an FCG}\label{alg:interpretation}
  \SetKwProg{Fn}{Function}{:}{}

@@

  \KwIn{$\mathcal{D}$---the set of domains of the input formula}
  \KwIn{$s$---the source node of the FCG}
  \KwData{$\functionNames = \emptyset$ (function names)}
  \KwOut{a list of function definitions}
  $\variableNames \gets \{\, \Omega \mapsto \newVariableName{} \mid \Omega \in \mathcal{D} \,\}$\;\label{line:varnames}
  \If{$s$ is a direct successor of a $\Reff$ node}{
    $(\expression, \functions) \gets \visit{$s$, $\variableNames$}$\;\label{line:initialcall}
    \Return{\functions}\;\label{line:initialcalltwo}
  }
  $f \gets \newFunctionName{$s$}$\;
  $(\expression, \functions) \gets \visit{$s$, $\variableNames$}$\;
  \Return{$\langle f(\langle \variableNames{$\Omega$} \mid \Omega \in \mathcal{D} \rangle) = \expression \rangle \mdoubleplus \functions$}\;
  \Fn{\visit{node $v$, variable names \variableNames}}{
    \If{$v$ is \textbf{not} a direct successor of a $\Reff$ node}{
      \Return{\actuallyVisit{$v$, \variableNames}}\;
    }
    $f \gets \newFunctionName{$v$}$\;
    $\newVariableNames \gets \variableNames$\;
    \For{$\Omega \in \mathfrak{D}(v)$ s.t. $\newVariableNames(\Omega)$ is non-atomic}{\label{line:nonone}
      $\newVariableNames(\Omega) \gets \newVariableName{}$\;\label{line:nontwo}
    }
    $(\expression, \functions) \gets \actuallyVisit{$v$, \newVariableNames}$\;\label{line:actuallyvisit}
    $\functionCall \gets f(\langle \variableNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    $\functionSignature \gets f(\langle \newVariableNames{$\Omega$} \mid \Omega \in \mathfrak{D}(v) \rangle)$\;
    \Return{$(\functionCall, \langle \functionSignature = \expression \rangle \mdoubleplus \functions)$}\;
  }
  \Fn{\actuallyVisit{node $v$, variable names \variableNames}}{
    \Switch{label of $v$}{
      \lCase{$\GDR(v')$}{\Return{\visit{$v'$, \variableNames}}}
      \uCase{$\CR_{\Omega \mapsto \Omega'}(v')$}{
        \Return{\visit{$v'$, $\variableNames \cup \{\, \Omega' \mapsto (\variableNames(\Omega) - 1) \,\}$}}\;\label{line:cr}
      }
      \uCase{$\Reff_{\rho}(v')$}{
        $\args \gets \langle \variableNames\circ\rho(\Omega) \mid \Omega \in \mathfrak{D}(v') \rangle$\;\label{line:args}
        \Return{$(\functionNames{$v'$}(\args), \langle \rangle)$}\;
      }
      \dots
    }
  }
\end{algorithm}

\begin{itemize}
  \item TODO: is maybe `parameter name' (or something similar) would be more
        appropriate than `variable name'?
  \item We call a variable name is \emph{atomic} if it is free of subtraction
        (e.g., $m$, $n$); otherwise it is \emph{non-atomic} (e.g., $m-1$,
        $n-l$).
  \item For each node $v$, let $\mathfrak{D}(v)$ denote the (precomputed) set of
        domains, the sizes of which we would use as parameters if we were to
        define a function that begins at $v$. This set is computed by
        iteratively setting
        $\mathfrak{D}(v) \gets \left(\bigcup_{u} \mathfrak{D}(u) \setminus I_{v}\right) \cup U_{v}$
        until convergence, where
        \begin{itemize}
          \item $u$ is any direct successor of $v$,
          \item $I_{v}$ is the set of domains \emph{introduced} at node $v$, and
          \item $U_{v}$ is the set of domains \emph{used} by $v$.
        \end{itemize}
  \item We assume a fixed total ordering of all domains names. In particular, in
        our running example we assume that $\Gamma$ goes before $\Delta$.
  \item $\mathsf{F}$ is a partial map from the set of FCG nodes to function
        names (e.g., $f$, $g$).
  \item \newVariableName just generates a unique identifier for a variable
  \item \newFunctionName does the same thing for function names and also records
        the new name in $\mathsf{F}$
  \item The first function in the list is the main one.
  \item Note that a node is a direct successor of a $\Reff$ node if and only if
        it has an in-degree greater than one.
  \item For simplicity, we refer to FCG nodes using their types.
  \item The algorithm consists of two main functions: \visit and \actuallyVisit.
        The former handles new function definitions, and the latter produces an
        algebraic interpretation of each node depending on its type. As there
        are many node types, \cref{alg:interpretation} only includes the ones
        pertinent to the contributions of this paper; see previous
        work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information about the other
        types.
  \item Given some node $v$ as input, both \visit and \actuallyVisit return a
        pair $(\mathsf{e}, \mathsf{funs})$. Here, $\mathsf{e}$ is the algebraic
        expression that should be used to evaluate $v$ or use it in another
        expression, and $\mathsf{funs}$ is a list of auxiliary functions that
        were created while constructing $\mathsf{e}$.
  \item TODO: $\mathfrak{D}$ must be a list. That's the ordering of domain names
        comes in, not in list comprehensions.
\end{itemize}

@@

\begin{figure}
  \centering
  \begin{tikzcd}
    \ar[d] & \\
    \text{\Cref{alg:interpretation}} \ar[d] \ar[u, swap, shift right=0.7ex, "{\left\langle f(m, n) = \sum_{l=0}^{n}[l<2]f(m-1,n-l) \right\rangle}"] & \\
    \visit{$\GDR$, $\Vone$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({f(m, n), \left\langle f(m, n) = \sum_{l=0}^{n}[l<2]f(m-1,n-l) \right\rangle}\right)"] & \\
    \actuallyVisit{$\GDR$, $\Vone$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({\sum_{l=0}^{n}[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\bigvee$, $\Vone$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({\sum_{l=0}^{n}[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\CR$, $\Vtwo$} \ar[d] \ar[u, swap, shift right=0.7ex, "\left({[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\land$, $\Vthree$} \ar[d] \ar[dr] \ar[u, swap, shift right=0.7ex, "\left({[l<2]f(m-1,n-l), \langle\rangle}\right)"] & \\
    \visit/\actuallyVisit{$\bot$, $\Vthree$} \ar[u, swap, shift right=0.7ex, "\left({[l<2], \langle\rangle}\right)"] & \visit/\actuallyVisit{$\Reff$, $\Vthree$} \ar[ul, swap, shift right=0.7ex, "\left({f(m-1, n-l), \langle\rangle}\right)"]
  \end{tikzcd}
  \caption{Function calls and their return values when \cref{alg:interpretation}
    is run on the FCG from \cref{fig:examplefcg}. When \protect\visit calls
    \protect\actuallyVisit and returns its output unchanged, the two function
    calls are shortened as a single call to
    $\protect\visit/\protect\actuallyVisit$. TODO: this last sentence should be
    explained in text.}\label{fig:functioncalls}
\end{figure}

\begin{example}\label{example:interpretation}
  Let us see how \cref{alg:interpretation} handles the FCG from
  \cref{fig:examplefcg}. Here, $\mathcal{D} = \{\, \Gamma, \Delta \,\}$. Let
  $\Vone \coloneqq \{\, \Gamma \mapsto m, \Delta \mapsto n \,\}$,
  $\Vtwo \coloneqq \Vone \cup \{\, \Delta^{\top} \mapsto l, \Delta^{\bot} \mapsto (n-l) \,\}$,
  and $\Vthree \coloneqq \Vtwo \cup \{\, \Gamma' \mapsto (m-1) \,\}$ denote
  versions of $\mathsf{V}$ at various points throughout the algorithm's
  execution. Here, $l$, $m$, and $n$ are all arbitrary variable names generated
  by \newVariableName. \Cref{line:varnames} initialises $\mathsf{V}$ to $\Vone$.

  \Cref{fig:functioncalls} shows the outline of function calls and their return
  values. Once \visit{$\GDR$, $\Vone$} is called on \cref{line:initialcall},
  \newFunctionName updates $\mathsf{F}$ to $\{\, \GDR \mapsto f \,\}$, where $f$
  is a new function name. There are no non-atomic names in $\Vone$, so
  \cref{line:nonone,line:nontwo} are skipped, and calls to
  \actuallyVisit{$\GDR$, $\Vone$} and
  $\visit/\actuallyVisit{$\bigvee$, $\Vone$}$ follow. Here, $\Vone$ becomes
  $\Vtwo$, i.e., the $\bigvee$ node introduces two domains $\Delta^{\top}$ and
  $\Delta^{\bot}$ that `partition' $\Delta$, i.e.,
  $\Vtwo(\Delta^{\top}) + \Vtwo(\Delta^{\bot}) = l + (n-l) = n = \Vtwo(\Delta)$.
  Similarly, \actuallyVisit{$\CR$, $\Vtwo$} on \cref{line:cr} adds $\Gamma'$,
  replacing $\Vtwo$ with $\Vthree$.

  Eventually, \actuallyVisit{$\bot$, $\Vthree$} is called. A contradiction node
  with clause $c$ as a parameter is interpreted as one if the clause has
  groundings and zero otherwise. In this case, the parameter is
  $c = (\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\})$
  (not shown in \cref{fig:examplefcg}). Equivalently,
  $\forall X, Y \in \Delta^{\top}\text{. }X \ne Y \Rightarrow \bot$, i.e.,
  $\forall X, Y \in \Delta^{\top}\text{. }X = Y$. This sentence is true if and
  only if $\Vthree(\Delta^{\top}) < 2$. Thus, we can use the Iverson bracket
  notation to write
  \[
    [l < 2] \coloneqq
    \begin{cases}
      1 & \text{if } l < 2 \\
      0 & \text{otherwise.}
    \end{cases}
  \]
  for the output expression of \actuallyVisit{$\bot$, $\Vthree$}.

  Next, \actuallyVisit{$\land$, $\Vthree$} calls
  $\visit/\actuallyVisit{$\Reff$, $\Vthree$}$. Since
  $\mathfrak{D}(\GDR) = \langle \Gamma, \Delta \rangle$, and
  $\begin{tikzcd} \langle\Gamma, \Delta\rangle \ar[r, mapsto, "\rho"] & \langle\Gamma', \Delta^{\bot}\rangle \ar[r, mapsto, "\Vthree"] & \langle m-1, n-l \rangle\end{tikzcd}$,
  \textsf{args} is set to $\langle m-1, n-l \rangle$ on \cref{line:args}, and
  the output expression becomes $f(m-1, n-l)$. The call to
  \actuallyVisit{$\land$, $\Vthree$} then returns the product of the two
  expressions $[l<2]f(m-1, n-l)$. Next, the call to \actuallyVisit{$\bigvee$,
    $\Vone$} prepends a summation and returns $\sum_{l=0}^{n}[l<2]f(m-1,n-l)$.
  The same expression (and an empty list of functions) is then returned to
  \cref{line:actuallyvisit} of the call to \visit{$\GDR$, $\Vone$}. Here, both
  \textsf{call} and \textsf{signature} are set to $f(m, n)$. As the definition
  of $f$ is the final answer and not part of some other algebraic expression,
  \cref{line:initialcall,line:initialcalltwo} discard the function call
  expression and return the definition of $f$.

  Therefore,
  \begin{align}
    f(m, n) &= \sum_{l = 0}^{n} \binom{n}{l} [l < 2] f(m-1, n-l)\nonumber \\
            &= f(m-1, n) + n f(m-1, n-1).\label{eq:solution}
  \end{align}
  To use this recursive function to compute the model count of the input formula
  for any domain sizes, one just needs to find the base cases $f(0, n)$ and
  $f(m, 0)$ for all $m, n \in \mathbb{N}_{0}$.
\end{example}

\section{Experimental Results}\label{sec:results}
% TODO: rename 'experimental results' -> 'complexity results' (not just the
% title of the section but review every mention of the word 'experimental')
% TODO: make it clear that I'm not doing this by hand anymore, but I'm still
% removing multiplications by one.
% TODO (here or somewhere else): emphasise that there are two parts: search
% (with some artificial limit) + inference (whose complexity we can easily
% determine)
% TODO: Must do one of these:
% 1. determine the complexity of the counting quantifiers algorithm when run on
% these instances
% 2. find instances that are not in C^2 and make the paper about them
% 3. run the counting quantifiers algorithm on these instances with increasing
% domain sizes, statistically estimating the degree of the polynomial and
% juxtaposing it with my complexity results shifted to overlap as much as
% possible

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: 'Asymptotic Complexity of Counting' -> 'Complexity of \mathcal{F}_\phi'
% (introduce \mathcal{F}_\phi in the introduction, together with the flow graph)

% TODO: remove the rows that say 'same as..', putting the explanation about them
% in text. And don't say 'same as', instead say that the formulas are in some
% sence equivalent (maybe in a stronger sense than the usual equivalence).

% TODO: emphasise that the reader should focus on the last two columns
% TODO: the last sentence of the caption could be confusing
% TODO: 'best known' sounds disappointing. Find a better alternative.

@@

% how the complexity of a solution is determined
Previous work often compares WFOMC algorithms by running them on a few instances
with increasing domain sizes and measuring
runtime~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/aaai/BroeckD12}.
However, we can do much better and identify the exact worst-case asymptotic
complexity of a solution produced by either \textsc{Crane} or \textsc{ForcLift}.
The asymptotic complexity of a solution can be determined by considering the
total number of arithmetic operations needed to follow the definitions of
constituent functions without recomputing the same quantity multiple times. In
particular, we assume that each function call and binomial coefficient is
computed at most once. We also assume that computing $\binom{n}{k}$ takes
$\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on the computation of each $f(i, j)$.
% TODO: look for a better place to put the 'In particular...' remark
